Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-04-19 document/ms.tex
adding derivation of maximum likelihood ide procedure

% Document specific variables.
\newcommand{\rate}{\ensuremath{\eta}}
\newcommand{\ratepars}{{\ensuremath{\bvec{\theta}}}}
\newcommand{\obs}[1]{\ensuremath{\hat{#1}}}
\newcommand{\radius}{\ensuremath{R}}
\newcommand{\period}{\ensuremath{T}}
\newcommand{\completeness}{{\ensuremath{P_\mathrm{c}}}}
\newcommand{\transitprob}{{\ensuremath{P_\mathrm{t}}}}
\newcommand{\entry}{\ensuremath{S}}
\newcommand{\catalog}{{\ensuremath{\bvec{\entry}}}}

@@

\newcommand{\bincenter}{{\ensuremath{\bvec{x}}}}
\newcommand{\binheight}{{\ensuremath{w}}}
\newcommand{\binheights}{{\ensuremath{\bvec{\binheight}}}}

@@

>>>>

% Document specific variables.
\newcommand{\rate}{\ensuremath{\Gamma}}
\newcommand{\ratepar}{{\ensuremath{\theta}}}
\newcommand{\ratepars}{{\ensuremath{\bvec{\ratepar}}}}
\newcommand{\obs}[1]{\ensuremath{\hat{#1}}}
\newcommand{\radius}{\ensuremath{R}}
\newcommand{\period}{\ensuremath{T}}
\newcommand{\completeness}{{\ensuremath{P_\mathrm{c}}}}
\newcommand{\transitprob}{{\ensuremath{P_\mathrm{t}}}}
\newcommand{\entry}{\ensuremath{w}}
\newcommand{\catalog}{{\ensuremath{\bvec{\entry}}}}

@@

\newcommand{\binarea}{{\ensuremath{\Delta}}}
\newcommand{\bincenter}{{\ensuremath{\bvec{x}}}}
\newcommand{\binheight}{{\ensuremath{w}}}
\newcommand{\binheights}{{\ensuremath{\bvec{\binheight}}}}

@@

\appendix

\section{Maximum likelihood estimators for noiseless catalogs}

Equation XXXX gives the likelihood of a catalog $\catalog =
\{\entry_k\}_{k=1}^K$, conditioned on a particular choice of occurence rate
parameters \ratepars:
\begin{eqnarray}
\ln p(\catalog\,|\,\ratepars) &=& \sum_{k=1}^K \ln \rate(\entry_k)
    - \int \rate(\entry) \dd\entry \quad.
\end{eqnarray}
If we model the occurence rate as a piecewise constant function
\begin{eqnarray}
\rate (\entry) &=& \left \{\begin{array}{ll}
\ratepar_1 & \mathrm{if\,} \entry \in \binarea_1, \\
\cdots & \cdots \\
\ratepar_N & \mathrm{if\,} \entry \in \binarea_N, \\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
where $\binarea_n$ is the volume of the bin $n$, the log likelihood becomes
\begin{eqnarray}
\ln p(\catalog\,|\,\ratepars) &=&
    \sum_{k=1}^K \sum_{n=1}^N \mathbf{1}[\entry_k \in \binarea_n]
        \ln\ratepar_n
    - \sum_{n=1}^N \ratepar_n\,\binarea_n \quad.
\end{eqnarray}
The maximum likelihood solution to this can be computed analytically
\begin{eqnarray}
\frac{\dd \ln p(\catalog\,|\,\ratepars)}{\dd \ratepar_m} &=&
    \frac{N_m}{\ratepar_m} - \binarea_m \\
\to \quad \ratepar_m &=& \frac{N_m}{\binarea_m}
\end{eqnarray}
where $N_m$ is the number of objects in the bin $m$.
This result is simply a \emph{normalized histogram} of the points.

Including the completeness function $\completeness(\entry)$ is as simple as
modeling the observable occurence rate as
\begin{eqnarray}
\rate (\entry) &=& \completeness(\entry) \, \left \{\begin{array}{ll}
\ratepar_1 & \mathrm{if\,} \entry \in \binarea_1, \\
\cdots & \cdots \\
\ratepar_N & \mathrm{if\,} \entry \in \binarea_N, \\
0 & \mathrm{otherwise}
\end{array}\right. \quad.
\end{eqnarray}
This leads to a similar analytic maximum likelihood solution
\begin{eqnarray}
\frac{\dd \ln p(\catalog\,|\,\ratepars)}{\dd \ratepar_m} &=&
    \frac{N_m}{\ratepar_m}
    - \int_{\binarea_m} \completeness(\entry)\dd\entry \\
\to \quad \ratepar_m &=& N_m \,
    \left [\int_{\binarea_m} \completeness(\entry)\dd\entry \right ]^{-1}
\end{eqnarray}
where the remaining integral is over the bin $m$.
This result is \emph{almost} the same as the inverse-detection-efficiency
procedure but each point is weighted by the bin-averaged efficiency instead of
the per-object efficiency.
